Quantum Hoare Type Theory

### 2020-10-30 ms-paper/ms-paper.tex
Update examples in 3.3, add entangle-disentangle example

\subsubsection{Hello Quantum World}
 \todo{TODO}
\leavevmode

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bit {\top /\ r=0}
    = do q <= init 0;
         return measure q
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bit {\top /\ r = 0}
-- P0: \top
    = do q <= init 0;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bit {\top}
    = do q <= init 0;
         q <= apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

In the case of quantum coin toss, the result can be in either 0 or 1, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.

Verification steps follow: \todo{TODO}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bit {\top}
-- P0: \top
    = do q <= init 0;
-- P1: P0 \o (q |-> |0\>)
         q <= apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode
\todo{TODO: syntax for applying multi-qubit gate}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
         = do qa <= init 0;
              qa <= apply H to qa;
              qb <= init 0;
              (qa, qb) <= apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.

\paragraph{Verification}
We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function. For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below. We have elided specific details about unitary application as they are still being worked out.

\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bit o* Bit) {\top /\ a = b}
              -- P0: \top
         = do qa <= init 0;
              -- P1: P0 \o (qa |-> |0\>)
              qa <= apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init 0;
              -- P3: P2 \o (qb |-> |0\>)
              (qa, qb) <= apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
\end{lstlisting}

@@

>>>>

\subsubsection{Hello Quantum World}
\leavevmode

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ r = false}
    = do q <= init false;
         r = meas q;
         return r
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ r = false}
         -- \top
    = do q <= init false;
         -- q =q |0\> /\ class(q)
         r = meas q;
         -- (q, e) =q |00\> /\ r = false -- could use a sugared version for measured
         return r
         -- r = false -- if no qubit left, should be able to switch to \top here
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
    = do q <= init false;
         apply H to q;
         r = meas q;
         return r
\end{lstlisting}
\end{minipage}

In the case of quantum coin toss, the result can be either true or false, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
         -- \top
    = do q <= init false;
         -- q =q |0\> /\ class(q)
         apply H to q;
         -- q =q |+\> /\ uniform(q)
         r = meas q;
         -- (q, e) =q |++\> -- perhaps we should be able to say uniform(r)?
         return r
         -- r = false \/ r = true -- or left unspecified
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
         = do qa <= init false;
              apply H to qa;
              qb <= init false;
              <= apply CNOT to (qa, qb);
              a = meas qa;
              b = meas qb;
              return (a, b)
\end{lstlisting}
\end{minipage}

Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.

\paragraph{Verification}
We would like to describe how QHTT lets us verify the correctness of the above program based on the specification provided as the type of the function. For illustration, we show the propositions annotated as comments as we step through the \texttt{testBell00} program below. We have elided specific details about unitary application as they are still being worked out.
\newpage
\begin{lstlisting}[language=QHaskell]
testBell00 : {\top} (a, b) : (Bool o* Bool) {\top /\ a = b}
              -- \top
         = do qa <= init false;
              -- qa =q |0\>, class(qa)
              apply H to qa;
              -- qa =q |+\>, uniform(qa)
              qb <= init false;
              -- qa =q |+\>, uniform(qa), qb =q |0\>, class(qb)
              apply CNOT to (qa, qb);
              -- (qa, qb) =q |++\>  -- can use a weaker but more general `EPR' here
              -- /\ uniform(qa, qb) -- this won't apply then
              a = meas qa;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true
              b = meas qb;
              --   (qa, e1) =q |00\>, (qb, e2) =q |00\>, a = false, b = false
              -- \/ (qa, e1) =q |11\>, (qb, e2) =q |11\>, a = true,  b = true
              return (a, b)
\end{lstlisting}

@@

\subsection{Entangle disentangle}
\todo{OPTIONAL:}

\begin{lstlisting}[language=QHaskell]
entangle-disentangle : {T} (a, b) : (Qbit o* Qbit) {a = b}
= do
    -- P0: T
    a <= init false;
    -- P1: a =q |0\>, class(a)
    b <= init false;
    -- P2: P1, b =q |0\>, class(b)
    apply H to a;
    -- P3: H_a.P2     =   a =q |+\>, b =q |0\>, uniform(a), class(b)
    apply CX to (a, b);
    -- P4: CX_ab.P3   =   (a, b) =q |++\>, uniform(a, b)
    apply CX to (a, b);
    -- P5: CX_ab.P4   =   (a =q |+\>, b =q |0\>) \/ (a =q |0\>, b =q |+\>)
    apply H to a;
    -- P6: H_a.P5     =   (a =q |0\>, b =q |0\>) \/ (a =q |+\>, b =q |+\>)
    return (a, b)
\end{lstlisting}
